---
title: "Activity Prediction"
author: "Stefan Faußer"
date: "11.05.2016"
output: html_document
---

# Summary

In this document, activities are predicted from accelerometers on the belt, forearm, arm, and dumbell of six participants. The data is from the Weight Lifting Exercises dataset from <http://groupware.les.inf.puc-rio.br/har>. The participipants were asked to perform five different fashions, ten repetitions, of the Unilateral Dumbbell Biceps Curl. These are: According to specification (A), throwing elbows to front (B), lifting halfway (C), lowering halfway (D), throwing hips to front (E).

Summarized, out of the evaluated models, the random forests performed best with an **out-of-sample accuracy** of more than $0.99$, followed by stochastic gradient boosting ($0.96$).

# Preprocessing

In this section, the data preprocessing steps are described.

```{r}
library(caret, quietly=TRUE)
library(corrplot, quietly=TRUE)
library(doMC, quietly=TRUE)
registerDoMC(cores = 6) # allow parallel processing with up to six cores
trainingFile <- "pml-training.csv"
trainData <- read.csv(trainingFile, na.strings = c("NA", ""))
```

**Predictor selection:** For reducing the complexity of the model, predictors are removed from the dataset according to the following critera:

- Irrelevant predictors. This includes: integer sequence (`X`), names (`user_name`), timestamps, windows, data columns: `1:7`.
- Predictors that have more than 90% of missing values (`NA`).
- Predictors with near zero variance.

```{r}
cols <- 1:7 # names, dates, timestamps
# get predictors that have more than 90% NAs
cols <- c(cols, which(colSums(is.na(trainData)) > nrow(trainData) * 0.9))
# get predictors with near zero variance
cols <- c(cols, nearZeroVar(trainData))
cols <- unique(cols)
trainData <- trainData[,-cols] # remove the predictors
print(sort(cols)) # outputs the removed predictors
```

## Collinearity

```{r}
corrs <- cor(trainData[,-ncol(trainData)])
corrplot(corrs, order = "hclust")
```

From above correlation plot we observe that there are some variables that are strongly correlated ($\geq 0.9$, red). These variables are removed for reducing the model complexity.

```{r}
cols <- findCorrelation(corrs, cutoff = 0.9)
trainData <- trainData[,-cols] # remove the predictors
print(sort(cols)) # outputs the removed predictors
```

# Model building

- The training dataset (`pml-training.csv`) is splitted into training ($0.6$ of the data) and testing using stratified sampling.

```{r}
set.seed(107)
inTrain <- createDataPartition(y = trainData$classe, p = 0.6, list = FALSE)
training <- trainData[inTrain,]
testing <- trainData[-inTrain,]
```

- The training dataset is used for building the model, $5$-fold cross validation with three repeats.
- The problem is a classification problem, for the outcome (`classe`) is categorical with five classes.
- Out of the popular classification methods, the two top performing methods random forests (`rf`) and stochastic gradient boosting (`gbm`) are evaluated.
- Note that feature normalization is unnecessary for decision trees (`gbm` and `svmPoly`).
- The evaluated criteria is the **accuracy** of the model. The error is, thus, $1 - accuracy$.

**Random forests:**

```{r, cache = TRUE}
set.seed(107)
rfFit <- train(classe ~.,
               data = training,
               method = "rf",
               trControl = trainControl(method = "repeatedcv", number = 5, repeats = 3))
rfCmTrain <- confusionMatrix(data = predict(rfFit, training), training$classe)
rfCmTrain
1 - as.vector(rfCmTrain$overall['Accuracy']) # in-sample error
```

**Stochastic gradient boosting with decision trees:**

```{r, cache = TRUE}
set.seed(107)
gbmFit <- train(classe ~.,                
                data = training,
                method = "gbm",
                verbose = FALSE,               
                trControl = trainControl(method = "repeatedcv", number = 5, repeats = 3))
gbmCmTrain <- confusionMatrix(data = predict(gbmFit, training), training$classe)
gbmCmTrain
1 - as.vector(gbmCmTrain$overall['Accuracy']) # in-sample error
```

- The random forest (`rf`) model performed best on the training dataset with an **in-sample accuracy** of $1.00$, followed by the stochastic gradient descent model (`gbm`, $0.97$). Note that these are optimistic values, see section *out-of-sample error* below.

# Out-of-sample error and model selection

- The testing dataset (see above) is used for retrieving the **out-of-sample errors**.
- The **out-of-sample errors** are expected to be higher than the **in-sample errors** (training errors), see section *Model building and selection* above, because the models were trained with the training data and are, thus, biased towards the training data.
- The model with the lowest **out-of-sample error** is considered to be the final model.

```{r, cache = TRUE}
rfCmTest <- confusionMatrix(data = predict(rfFit, testing), testing$classe) # random forests
rfCmTest
1 - as.vector(rfCmTest$overall['Accuracy']) # out-of-sample error
gbmCmTest <- confusionMatrix(data = predict(gbmFit, testing), testing$classe) # stochastic gradient boosting
gbmCmTest
1 - as.vector(gbmCmTest$overall['Accuracy']) # out-of-sample error
```

- The random forest (`rf`) model has an **out-of-sample accuracy** of more than $0.99$ on the testing dataset, followed by the stochastic gradient descent model (`gbm`, $0.96$).
